Embedded Domain Specific Languages inside Functional Languages

### 2014-03-13 paper.tex
Basic DESDL intro + first example

@@

But not everything is a simulation. What if we wanted to use an EDSL to express
something that we want to run somewhere else, not inside the Haskell system?
Can we use Lava to generate circuits, and run them on FPGAs? Can we use EDSL
to generate code for embedded processors, or GPUs? Such an ability, to 
synthesis external solutions, would be extremely useful. And we can extend
the EDSL idiom to do so, with significant caveats. The remainder of this
paper is about how we capture and offshore work from inside a EDSL, what
this capability can be used for, and what the limitations are.

@@

As can be seen, EDSLs are simply a way of thinking about a library
of provided functions, often called combinators, because they combine
their arguments into terms inside the DSL language.

Compositionally
Combining elements into more complex ones should be easy and natural.
Abstraction
The user shouldn’t have to know (or be allowed to exploit) the underlying implementation of your types.

(((MORE))

There are two flavors of EDSLs.
The most common flavor are
{\em DSLs that use a shallow embedding\/}, where values are computed with directly.
The result of a computation in a shallow DSL is a value.
But there is another class of DSLs, 
specifically {\em DSLs that use a deep embedding\/} build an abstract syntax tree~\cite{Elliott:03:CompileDSEL-JFP}.
The result of a computation inside a deep DSL
is a structure, not a value, and this structure can be used to compute a value,
or be cross-compiled before being evaluated.

This is an important distinction. Historically, EDSLs have been shallow,
and as such are simply a ways of structuring an API to a library. Deep EDSLs
have the ability to {\em stage\/} code, that is executing a program
can generate another program, much like the well-known YACC DSL,
but at the cost of significantly restricting what forms of the DSL can
generate valid output from.
%% That sentence is a bit garbled.

The unifying theme is that embedded DSLs can
be pragmatic, productive and useful.

\section{Deep Embeddings of DSLs}

A deeply embedded DSL is a DSL than can have ???.
To make this concrete, consider a DSL for arithmetic,
with addition, subtraction, multiplication, and constants.
We want to be able to {\bf run\/} this DSL. For a shallow
embedding this is trival - we just use the built in 
arithmetic. A deep embedding is where things get
interesting. Consider a data-type for our arithmetic.

Now we overload the arithemtic to use this E datatype;
in Haskell Num is our overloading for integeral arithmetic.

\begin{Code}
instance Num E where
  e1 + e2 = Add e1 e2
  ...

With deep embeddings, it is common to also
write a run function, that computes the structure
of a catured computation, given a simple value.
\begin{Code}
run :: E -> Int
run (...)
\end{Code}

>>>>

Not everything is a simulation. What if we wanted to use an EDSL to express
something that we want to run somewhere else, not inside the Haskell system?
Can we use Lava to generate circuits, and run them on FPGAs? Can we use EDSL
to generate code for embedded processors, or GPUs? Such an ability, to 
synthesis external solutions, would be extremely useful. We can extend
the EDSL idiom to do so, with significant caveats. The remainder of this
paper is about how we capture and offshore work from inside a EDSL, what
this capability can be used for, and what the limitations are.

@@

EDSLs are simply a way of thinking about a library
of provided functions, often called combinators, because they combine
their arguments into terms inside the DSL language and types. In our
Lava example above, the \verb|register| combinator takes an initial
value, and a incoming stream of values, and gives the new stream,
delayed by a single cycle, with the initial value occupying the initial
cycle. Critically, \verb|register| is compositional; it combines
smaller parts of the DSL to make larger solutions. If a DSL
follows this composability carefully by design, an alternative implementation
is possible.

The most common flavor of DSL is one  that uses a so-called shallow embedding, where values are computed with directly.
The result of a computation in a shallow DSL is a value. All the examples so far are shallow.
But there is another class of DSLs, 
specifically {\em DSLs that use a deep embedding\/} build an abstract syntax tree~\cite{Elliott:03:CompileDSEL-JFP}.
The result of a computation inside a Deeply Embedded DSL (Deep EDSL)
is a structure, not a value, and this structure can be used to compute a value,
or be cross-compiled before being evaluated. Such ``Deep'' EDSLs
follow the composability mantra pedantically, by design and mandate.

Historically, EDSLs have been shallow; simply a way of structuring an API to a library. 
However, Deep EDSLs have the ability to {\em stage\/} code, that is executing a program
can generate another program, much like the well-known YACC DSL,
but at the cost of significantly restricting what forms of the DSL can
generate valid output from.
There is a growing number of Deep EDSLs, and a body of research around their
form and limitations.
The unifying theme is that Deep EDSLs can be pragmatic, productive and useful.

In this section, we will investigate the basic structure of Deep EDSL compared to Shallow EDSLs,
and look at three pragmatic tricks for improving the usefulness of Deep EDSLs.

\subsection{Building a Deep EDSL}

A deeply embedded DSL is a DSL that exposes its own structure.
Rather than functions operating on a datastructure (a shallow DSL),
a deep DSL build a structure, and allows some secondary agent to
provide interpretation of this structures.
To make this idea concrete, consider a DSL for arithmetic,
with addition, subtraction, multiplication, and constants.
We want to be able to {\bf run\/} this DSL. For a shallow
embedding this is trivial - we just use the built in 
arithmetic. A deep embedding is where things get
interesting. Consider a data-type for our arithmetic.

Now we overload the arithmetic to use this E datatype;
in Haskell Num is our overloading for integral arithmetic.

\begin{Code}
instance Num E where
  fromInteger n = Lit n
  e1 + e2 = Add e1 e2
  e1 - e2 = Sub e1 e2
  e1 * e2 = Mul e1 e2
\end{Code}

This profound, and the key idea that makes deep embeddings work. We can written
an expression, and extracted a tree of {\em what\/} to do, not a direct result.
With deep embeddings, it is common to also
write a run function, that computes the structure
of a captured computation, given a simple value.
\begin{Code}
run :: E -> Int
run (Lit n) = n
run (Add a b) = run a + run b
run (Sub a b) = run a - run b
run (Mul a b) = run a * run b
\end{Code}

@@
